### 2010-07-26 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: we can even afford some fanciness on that line
Desc@ICFP: remove odd \newpage break
gaol: minor last-minute tweakage

Just as in the universe of descriptions, this construction comes with
an obvious \return and a substitution operation, the \bind. Its
definition is the following:
%
\[\stk{
\begin{array}{@{}ll}
\substI : & _{\PI{\V{I}}{\Set}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
          & (\V{X} \DotTo 
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
\end{array} \\
\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
\qquad    \cataI_{\V{I}}\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
                      (\muide{\V{Y}}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
                      \V{i}\:
                      \V{t} 
}\]
% 
Where  $\F{applyI}$ is defined as follow:
%
\[\stk{
\begin{array}{@{}ll}
\F{applyI} : & _{\PI{\V{I}}{\Set}}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
          & (\V{X} \DotTo \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                         {\V{I}}
                                         {\V{X}}}}
                    {\V{I}}
                    {\muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
            \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
\end{array}
}\]

@@

In the previous section, we have presented a language of closed
arithmetic expressions. Using the free monad construction, we are
going to extend this construction to open terms. An open term is
defined with respect to a context, represented by a snoc-list of
types:
%
\[
\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
\Ctxt           & \Set \\
\SYMBCtxtEmpty  & \Ctxt \\
\SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
\end{array}
\]
%
An environment realises the context, packing a value for each type:
%
\[
\stk{
\SYMBEnv : \Ctxt \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Env{\CtxtEmpty}                & \Unit \\
\Env{(\CtxtSnoc{\V{G}}{\V{S}})} & \TIMES{\Env{\V{G}}}{\Val{\V{S}}}
\end{array}
}\]
%
In this setting, we define $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \Ctxt \To \Ty \To \Set \\
\begin{array}{@{}ll@{\:\:\mapsto\:\:}l}
\Var{\CtxtEmpty}{& \V{T}}                & 
    \Void \\
\Var{(\CtxtSnoc{\V{G}}{\V{S}})}{& \V{T}} & 
    \SUM{(\Var{\V{G}}{\V{T}})}{(\V{S} \PropEq \V{T})}
\end{array}
}\]
%
While $\SYMBVal$ maps the type to the corresponding host type,
$\SYMBVar$ indexes a value in the context, obtaining a proof that the
types match. The $\SYMBlookup$ function precisely follow this
semantics:
%
\[\stk{
\SYMBlookup : \PI{\V{G}}{\Ctxt} 
          \Env{\V{G}} \To 
          \PI{\V{T}}{\Ty} 
          \Var{\V{G}}{\V{T}} \To
          \Val{\V{T}} \\
\begin{array}{@{}l@{}l@{}l@{}l@{}lll}
\lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
\lookup{& (\CtxtSnoc{\V{G}}{\V{S}})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumLeft{\V{x}})} & \mapsto & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
\end{array}
}\]

@@

Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
\V{G}\), we obtain the language of open terms in a context \(\V{G}\):
%
\[
\openTerm{\V{G}} \mapsto \FreeIMonad{\HExprD}{\Ty}{(\SYMBVar\:\V{G})}
\]
%
In this setting, the language of closed terms corresponds to the free
monad assigning an empty set of values to variables
%
\[
\closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
\quad
\mbox{where}
\quad
\stk{
\SYMBEmpty : \Ty \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Empty{\_}   & \Zero \\
\end{array}
}\]
%
Allowing variables from an empty set is much like forbidding variables,
so \(\closeTerm\) and \(\HExprD\) describe isomorphic
datatypes. Correspondingly, you can update an \(\HExprD\) to a
\(\closeTerm\):
%
\[\stk{
\update : \muide{\Ty}{\HExprD} \DotTo \muide{\Ty}{\closeTerm} \\
\begin{array}{@{}l@{}l}
\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & (\toIDesc{\HExprD})\:
                                                  (\muide{\Ty}{\closeTerm})\: \\
                                                & (\LAM{\_} \LAM{\pair{\V{tag}}{\V{tm}}{}} \Con{\pair{\Su{\V{tag}}}{\V{tm}}{}})\:
                                                  \V{ty}\:
                                                  \V{tm}
\end{array}
}\]
% 
The other direction of the isomorphism is straightforward, the
$\SYMBDVar$ case being impossible. Therefore, we are entitled to
reuse the $\evalH$ function to define the semantics of $\closeTerm$.

@@

An interesting instance of free monad is $\SYMBIDesc$ itself. Indeed,
$\SYMBDVar$ is nothing but the \return. The remaining constructors form
the carrier functor, trivially indexed by $\Unit$. The carrier functor
is described as follow:
%
\[\stk{
\IDescFreeD : \ATagIDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
                                \sqr{\SYMBDConst\:\:
                                     \DProd\:\:
                                     \SYMBDSigma\:\:
                                     \SYMBDPi} \red{,}\\
                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}}\end{array}}
\end{array}
}\]
%
Then, we get $\IDesc$ by building its free monad:
%
\[\stk{
\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}\sqr{}\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
}\]

@@

\newpage

>>>>

Just as in the universe of descriptions, this construction comes with
an obvious \return and a substitution operation, the \bind. Its
definition is the following:
%
\[\stk{
\begin{array}{@{}ll}
\substI : & _{\PI{\V{I}}{\Set}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
          & (\V{X} \DotTo 
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
\end{array} \\
\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
\qquad    \cataI_{\V{I}}\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
                      (\muide{\V{Y}}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
                      \V{i}\:
                      \V{t} 
}\]
% 
where  $\F{applyI}$ is defined as follows:
%
\[\stk{
\begin{array}{@{}ll}
\F{applyI} : & _{\PI{\V{I}}{\Set}}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
          & (\V{X} \DotTo \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                         {\V{I}}
                                         {\V{X}}}}
                    {\V{I}}
                    {\muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
            \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
\end{array}
}\]

@@

In the previous section, we presented a language of closed
arithmetic expressions. Using the free monad construction, we are
going to extend this construction to open terms. An open term is
defined with respect to a context, represented by a snoc-list of
types:
%
\[
\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
\Ctxt           & \Set \\
\SYMBCtxtEmpty  & \Ctxt \\
\SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
\end{array}
\]
%
An environment realises the context, packing a value for each type:
%
\[
\stk{
\SYMBEnv : \Ctxt \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Env{\CtxtEmpty}                & \Unit \\
\Env{(\CtxtSnoc{\V{G}}{\V{S}})} & \TIMES{\Env{\V{G}}}{\Val{\V{S}}}
\end{array}
}\]
%
In this setting, we define type variables, $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \Ctxt \To \Ty \To \Set \\
\begin{array}{@{}ll@{\:\:\mapsto\:\:}l}
\Var{\CtxtEmpty}{& \V{T}}                & 
    \Void \\
\Var{(\CtxtSnoc{\V{G}}{\V{S}})}{& \V{T}} & 
    \SUM{(\Var{\V{G}}{\V{T}})}{(\V{S} \PropEq \V{T})}
\end{array}
}\]
%
While $\SYMBVal$ maps the type to the corresponding host type,
$\SYMBVar$ indexes a value in the context, obtaining a proof that the
types match. The $\SYMBlookup$ function precisely follow this
semantics:
%
\[\stk{
\SYMBlookup : \PI{\V{G}}{\Ctxt} 
          \Env{\V{G}} \To 
          \PI{\V{T}}{\Ty} 
          \Var{\V{G}}{\V{T}} \To
          \Val{\V{T}} \\
\begin{array}{@{}l@{}l@{}l@{}l@{}lll}
\lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
\lookup{& (\CtxtSnoc{\V{G}}{\V{S}})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumLeft{\V{x}})} & \mapsto & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
\end{array}
}\]

@@

Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
\V{G}\), we obtain the language of open terms in a context \(\V{G}\):
%
\[
\openTerm{\V{G}} \mapsto \FreeIMonad{\HExprD}{\Ty}{(\SYMBVar\:\V{G})}
\]
%
In this setting, the language of closed terms corresponds to the free
monad assigning an empty set of values to variables
%
\[
\closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
\quad
\mbox{where}
\quad
\stk{
\SYMBEmpty : \Ty \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Empty{\_}   & \Zero \\
\end{array}
}\]
%
Allowing variables from an empty set is much like forbidding variables,
so \(\closeTerm\) and \(\HExprD\) describe isomorphic
datatypes. Correspondingly, you can update an old \(\HExprD\) to a shiny
\(\closeTerm\):
%
\[\stk{
\update : \muide{\Ty}{\HExprD} \DotTo \muide{\Ty}{\closeTerm} \\
\begin{array}{@{}l@{}l}
\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & (\toIDesc{\HExprD})\:
                                                  (\muide{\Ty}{\closeTerm})\: \\
                                                & (\LAM{\_} \LAM{\pair{\V{tag}}{\V{tm}}{}} \Con{\pair{\Su{\V{tag}}}{\V{tm}}{}})\:
                                                  \V{ty}\:
                                                  \V{tm}
\end{array}
}\]
% 
The other direction of the isomorphism is straightforward, the
$\SYMBDVar$ case being impossible. Therefore, we are entitled to
reuse the $\evalH$ function to define the semantics of $\closeTerm$.

@@

An interesting instance of free monad is $\SYMBIDesc$ itself. Indeed,
$\SYMBDVar$ is nothing but the \return. The remaining constructors form
the carrier functor, trivially indexed by $\Unit$. The signature functor
is described as follow:
%
\[\stk{
\IDescFreeD : \ATagIDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
                                \sqr{\SYMBDConst\:\:
                                     \DProd\:\:
                                     \SYMBDSigma\:\:
                                     \SYMBDPi} \red{,}\\
                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}}\end{array}}
\end{array}
}\]
%
We get $\IDesc{\V{I}}$ by extending the signature with variables from \(\V{I}\):
%
\[\stk{
\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}\sqr{}\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
}\]

@@

%\newpage

### 2010-07-26 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: helplessly stare at whitespaces and parenthesis
gaol: minor last-minute tweakage

Context validity ensures that variables inhabit well-formed
sets.
%
\[
%% Empty context validity
\Axiom{\vdash \Valid}
\qquad
%% Extend context
\Rule{\G       \vdash \Type{\M{S}}}
     {\G ; \xS \vdash \Valid}\;\x\not\in\G
\]
%
\begin{wstructure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Harper-Pollack, Luo, Courant
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{wstructure}
%
The basic typing rules
% (Fig.~\ref{fig:typing-judgements})
for tuples and functions are also standard, save that we locally adopt
\(\Set:\Set\) for presentational purposes. Usual techniques to resolve
this \emph{typical ambiguity} apply~\cite{harper:implicit-universe,
  luo:utt, courant:explicit-universe}. A formal presentation of a
stratified system is beyond the scope of this paper and is subject of
ongoing work.
%%  putting presentation before paradox~\cite{girard:set-in-set}.
%% The usual remedies apply, \emph{stratifying}
%% \(\Set\)~\cite{harper:implicit-universe, luo:utt,
%%   courant:explicit-universe}.
%%
\input{figure_typing_judgements}

@@

It is time for our first example of a \emph{universe}. You might
want to offer a choice of named constructors in your datatypes, we shall
equip you with sets of tags to choose from. Our plan
is to implement (by extending the theory, or by encoding) the signature
%
\[
  \Type{\EnumU}\qquad \Type{\EnumT{(\Bhab{\M{E}}{\EnumU})}}
\]
%
where some value \(E:\EnumU\) in the `enumeration universe' describes
a type of tag choices \(\EnumT{E}\). We shall need
some tags---valid identifiers, marked to indicate that
they are data, not variables scoped and substitutable---so we hardwire these
rules:
%
\[
%% UId
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Type{\UId}}
\qquad
%% Tag
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\Tag{\V{s}}}{\UId}}\;\V{s}\: \mbox{a valid identifier}
\]
%
Let us describe enumerations as lists of tags, with signature:
%
\[
\Bhab{\NilE}{\EnumU}\qquad
\Bhab{\ConsE{(\Bhab{\M{t}}{\UId})}{(\Bhab{\M{E}}{\EnumU})}}{\EnumU}
\]
%
What are the \emph{values} in \(\EnumT{E}\)? Formally, we represent
the choice of a tag as a numerical index into \(E\), via new rules:
%
\[
%% Ze
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\Ze}{\EnumT{(\ConsE{\M{t}}{\M{E}})}}} 
\qquad
%% Su
\Rule{\Gamma \vdash \Bhab{\M{n}}{\EnumT{\M{E}}}}
     {\Gamma \vdash \Bhab{\Su{\M{n}}}{\EnumT{(\ConsE{\M{t}}{\M{E}})}}}
\]
%
However, we expect that in practice, you might rather refer to these
values \emph{by tag}, and we shall ensure that this is possible in due course.

@@

We first define the `small product' $\SYMBspi$ operator:
%
\[\stk{
%% spi
\spi{}{}: \PITEL{\V{E}}{\EnumU}\PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To \Set \\
\begin{array}{@{}l@{\:}l@{\:\mapsto\:\:}l}
\spi{\NilE}{& \V{P}}        & \Unit \\
\spi{(\ConsE{\V{t}}{\V{E}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{E}}{(\LAM{\V{x}} \V{P}\: (\Su{\V{x}}))}}
\end{array}
}\]
%
This builds a right-nested
tuple type, demanding an object of $\V{P}\:\V{i}$ for each $\V{i}$ in the given finite
domain. We can see these tuples as `jump tables' tabulating dependently
typed functions from the domain. We give this functional
interpretation---the eliminator we need---by the
$\SYMBswitch$ operator, which, unsurprisingly, iterates projection:
%
\[\stk{
%% switch
\begin{array}{@{}ll}
\SYMBswitch : \PITEL{\V{E}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To
              \spi{\V{E}}{\V{P}} \To
               \PITEL{\V{x}}{\EnumT{\V{E}}} \To \V{P}\: \x
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\switch{(\ConsE{\V{t}}{\V{E}})}{\V{P}}{\V{b}}{\Ze}      & \fst{\V{b}} \\
\switch{(\ConsE{\V{t}}{\V{E}})}{\V{P}}{\V{b}}{(\Su{\V{x}})} & \switch{\V{E}}{(\LAM{\V{x}} \V{P}
  (\Su{\V{x}}))}{(\snd{\V{b}})}{\V{x}}
\end{array}
}\]

>>>>

Context validity ensures that variables inhabit well-formed
sets.
%
\[
%% Empty context validity
\Axiom{\vdash \Valid}
\qquad
%% Extend context
\Rule{\G       \vdash \Type{\M{S}}}
     {\G ; \xS \vdash \Valid}\;\x\not\in\G
\]
%
\begin{wstructure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Harper-Pollack, Luo, Courant
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{wstructure}
%
The basic typing rules
% (Fig.~\ref{fig:typing-judgements})
for tuples and functions are also standard, save that we locally adopt
\(\Set:\Set\) for presentational purposes. Usual techniques to resolve
this \emph{typical ambiguity} apply~\cite{harper:implicit-universe,
  luo:utt, courant:explicit-universe}. A formal treatment of
stratification for our system is a matter of
ongoing work.
%%  putting presentation before paradox~\cite{girard:set-in-set}.
%% The usual remedies apply, \emph{stratifying}
%% \(\Set\)~\cite{harper:implicit-universe, luo:utt,
%%   courant:explicit-universe}.
%%
\input{figure_typing_judgements}

@@

It is time for our first example of a \emph{universe}. You might
want to offer a choice of named constructors in your datatypes: we shall
equip you with sets of tags to choose from. Our plan
is to implement (by extending the theory, or by encoding) the signature
%
\[
  \Type{\EnumU}\qquad \Type{\EnumT{(\Bhab{\M{E}}{\EnumU})}}
\]
%
where some value \(E:\EnumU\) in the `enumeration universe' describes
a type of tag choices \(\EnumT{E}\). We shall need
some tags---valid identifiers, marked to indicate that
they are data, not variables scoped and substitutable---so we hardwire these
rules:
%
\[
%% UId
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Type{\UId}}
\qquad
%% Tag
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\Tag{\V{s}}}{\UId}}\;\V{s}\: \mbox{a valid identifier}
\]
%
Let us describe enumerations as lists of tags, with signature:
%
\[
\Bhab{\NilE}{\EnumU}\qquad
\Bhab{\ConsE{(\Bhab{\M{t}}{\UId})}{(\Bhab{\M{E}}{\EnumU})}}{\EnumU}
\]
%
What are the \emph{values} in \(\EnumT{E}\)? Formally, we represent
the choice of a tag as a numerical index into \(E\), via new rules:
%
\[
%% Ze
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\Ze}{\EnumT{(\ConsE{\M{t}}{\M{E}})}}} 
\qquad
%% Su
\Rule{\Gamma \vdash \Bhab{\M{n}}{\EnumT{\M{E}}}}
     {\Gamma \vdash \Bhab{\Su{\M{n}}}{\EnumT{(\ConsE{\M{t}}{\M{E}})}}}
\]
%
However, we expect that in practice, you might rather refer to these
values \emph{by tag}, and we shall ensure that this is possible in due course.

@@

We first define the `small product' $\SYMBspi$ operator:
%
\[\stk{
%% spi
\spi{}{}: \PITEL{\V{E}}{\EnumU}\PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To \Set \\
\begin{array}{@{}l@{\:}l@{\;\;\mapsto\;\;}l}
\spi{\NilE}{& \V{P}}        & \Unit \\
\spi{(\ConsE{\V{t}}{\V{E}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{E}}{\LAM{\V{x}} \V{P}\: (\Su{\V{x}})}}
\end{array}
}\]
%
This builds a right-nested
tuple type, packing a $\V{P}\:\V{i}$ value for each $\V{i}$ in the given
domain. The step case exposes our notational convention that
binders scope rightwards as far as possible. These tuples are `jump tables', tabulating dependently
typed functions.
We give this functional
interpretation---the eliminator we need---by the
$\SYMBswitch$ operator, which, unsurprisingly, iterates projection:
%
\[\stk{
%% switch
\begin{array}{@{}ll}
\SYMBswitch : \PITEL{\V{E}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To
              \spi{\V{E}}{\V{P}} \To
               \PITEL{\V{x}}{\EnumT{\V{E}}} \To \V{P}\: \x
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\switch{(\ConsE{\V{t}}{\V{E}})}{\V{P}}{\V{b}}{\Ze}      & \fst{\V{b}} \\
\switch{(\ConsE{\V{t}}{\V{E}})}{\V{P}}{\V{b}}{(\Su{\V{x}})} & \switch{\V{E}}{(\LAM{\V{x}} \V{P}
  (\Su{\V{x}}))}{(\snd{\V{b}})}{\V{x}}
\end{array}
}\]
